---
title: "Milestone Report"
author: "Nick Orka"
date: "4/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Synopsis

This is the Milestone Report for week 2 of the Coursera Data Science Capstone
project.

The objective of this report is to develop an understanding of the various
statistical properties of the data set that can later be used when building the
prediction model for the final data product - the Shiny application. Using 
exploratory data analysis, this report describes the major features of the
training data and then summarizes my plans for creating the predictive model.

The model will be trained using a unified document corpus compiled from the
following three sources of text data:

1. Blogs
1. News
1. Twitter

The provided text data are provided in four different languages. This project
will only focus on the English corpora.

## Environment Setup

Prepare the session by loading initial packages and clearing the global
workspace (including hidden objects).


```{r prepare_data, echo=FALSE, message=FALSE, error=FALSE}
dataUrl <- "https://d396qusza40orc.cloudfront.net/dsscapstone/dataset/Coursera-SwiftKey.zip"
zipFile <- "data/Coursera-SwiftKey.zip"

if (!file.exists('data')) {
    dir.create('data')
}

if (!file.exists(zipFile)) {
    download.file(dataUrl, zipFile, method = "curl")
}
if (!file.exists("data/final/en_US")) {
    unzip(zipFile, exdir = "data")
}
usFiles <- "data/final/en_US/"
fileTypes  <- c("blogs", "news", "twitter")
files <- sapply(fileTypes, function(type) {
    c <- file(paste0(usFiles, "en_US.", type, ".txt"), "r") 
    f <- readLines(c, encoding = "UTF-8", skipNul = T)
    close(c)
    f})
```

```{r files_stat, echo=F, cache=T}
lineChars <- sapply(files, function(l) {nchar(l)})
lineWords <- sapply(files, function(l) {sapply(gregexpr("[[:alpha:]]+", l), function(x) sum(x > 0))})
fileStats <- data.frame(
    Size_MB = round(sapply(lineChars, sum) / 1024 ^ 2),
    nWords = sapply(lineWords, sum),
    nLines = sapply(files, length),
    minWords = sapply(lineWords, min),
    meanWords = round(sapply(lineWords, mean)),
    maxWords = sapply(lineWords, max)
)

sampleSize <- 0.01
```


## Basic Data Summary

Prior to building the unified document corpus and cleaning the data, a basic
summary of the three text corpora is being provided which includes file
sizes, number of lines, number of characters, and number of words for each
source file. Also included are basic statistics on the number of words per line
(min, mean, and max).

### Initial Data Summary

```{r stat_table, echo=F}
kable(fileStats)
```

An initial investigation of the data shows that on average, each text corpora
has a relatively low number of words per line. Blogs tend to have more words
per line, followed by news and then twitter which has the least words per line.
The lower number of words per line for the Twitter data is expected given that a
tweet is limited to a certain number of characters. Even when Twitter doubled
its character count from 140 to 280 characters in 2017, research shows that only
1% of tweets hit the 280-character limit, and only 12% of tweets are longer than
140 characters. Perhaps after so many years, users were simply trained to the
140-character limit.

Another important observation in this initial investigation shows that the text
files are fairly large. To improve processing time, a sample size of
`r round(sampleSize*100)`% will be obtained from all three data sets and then
combined into a unified document corpus for subsequent analyses later in this
report as part of preparing the data.

### Histogram of Words per Line

```{r hist_words, echo=F, message=F, warning=F}
library(ggplot2)
library(gridExtra)

p1 <- qplot(lineWords$blogs, geom = "histogram", main = "Blogs", xlab = "Words Per Line", binwidth = 5)
p2 <- qplot(lineWords$news, geom = "histogram", main = "News", xlab = "Words Per Line", binwidth = 5)
p3 <- qplot(lineWords$twitter, geom = "histogram", main = "Tweets", xlab = "Words Per Line", binwidth = 5)

pList = list(p1, p2, p3)

do.call(grid.arrange, c(pList, list(ncol = 1)))
rm(p1, p2, p3, pList)
gc()
```

```{r sample_data, echo = FALSE, message=F, warning=F}
# set seed for reproducability
set.seed(1973)

# sample all three data sets
# remove all non-English characters from the sampled data
samples <- sapply(files, function(t) {
    iconv(sample(t, length(t) * sampleSize, replace = F), "latin1", "ASCII", sub = "")
})

# combine all three data sets into a single data set and write to disk
sampleFile <- "data/final/en_US/en_US.sample.txt"
con <- file(sampleFile, open = "w")
samples <- unlist(samples, use.names = F)
writeLines(samples, con)
close(con)

# get number of lines and words from the sample data set
numLines <- length(samples);
numWords <- sum(sapply(samples, function(l) {sapply(gregexpr("[[:alpha:]]+", l), function(x) sum(x > 0))}))

# remove variables no longer needed to free up memory
rm(files, lineChars, lineWords)
gc()
```


Prior to performing exploratory data analysis, the three data sets will be
sampled at `r round(sampleSize*100)`% to improve performance. All non-English
characters will be removed from the subset of data and then combined into a
single data set. The combined sample data set will be written to disk which
contains
`r format(round(as.numeric(numLines)), big.mark = ",")` 
lines and
`r format(round(as.numeric(numWords)), big.mark = ",")` 
words.

The next step is to create a corpus from the sampled data set. A custom
function named `buildCorpus` will be employed to perform the following
transformation steps for each document:

1. Remove URL, Twitter handles and email patterns by converting them to spaces using a custom content transformer
1. Convert all words to lowercase
1. Remove common English stop words
1. Remove punctuation marks
1. Remove numbers
1. Trim whitespace
1. Remove profanity
1. Convert to plain text documents

The corpus will then be written to disk in two formats: a serialized R object
in RDS format and as a text file. Finally, the first 10 documents (lines) from
the corpus will be displayed.

```{r clean_data, echo = FALSE, message=F, warning=F}
library(tm)

# download bad words file
badWordsURL <- "http://www.idevelopment.info/data/DataScience/uploads/full-list-of-bad-words_text-file_2018_07_30.zip"
badWordsZip <- "data/final/en_US/full-list-of-bad-words_text-file_2018_07_30.zip"
badWordsFile <- "data/final/en_US/full-list-of-bad-words_text-file_2018_07_30.txt"
if (!file.exists('data')) {
    dir.create('data')
}
if (!file.exists(badWordsFile)) {
    download.file(badWordsURL, badWordsZip, method = "curl")
    unzip(badWordsZip, exdir = "data/final/en_US/")
}

buildCorpus <- function (dataSet) {
    docs <- VCorpus(VectorSource(dataSet))
    toSpace <- content_transformer(function(x, pattern) gsub(pattern, " ", x))
    
    # remove URL, Twitter handles and email patterns
    docs <- tm_map(docs, toSpace, "(f|ht)tp(s?)://(.*)[.][a-z]+")
    docs <- tm_map(docs, toSpace, "@[^\\s]+")
    docs <- tm_map(docs, toSpace, "\\b[A-Z a-z 0-9._ - ]*[@](.*?)[.]{1,3} \\b")
    
    # remove profane words from the sample data set
    con <- file(badWordsFile, open = "r")
    profanity <- readLines(con, encoding = "UTF-8", skipNul = TRUE)
    close(con)
    profanity <- tail(profanity, length(profanity) - 14)
    profanity <- iconv(profanity, "latin1", "ASCII", sub = "")
    docs <- tm_map(docs, removeWords, profanity)
    
    docs <- tm_map(docs, tolower)
    docs <- tm_map(docs, removeWords, stopwords("english"))
    docs <- tm_map(docs, removePunctuation)
    docs <- tm_map(docs, removeNumbers)
    docs <- tm_map(docs, stripWhitespace)
    docs <- tm_map(docs, PlainTextDocument)
    return(docs)
}

# build the corpus and write to disk (RDS)
corpus <- buildCorpus(samples)
saveRDS(corpus, file = "data/final/en_US/en_US.corpus.rds")

# convert corpus to a dataframe and write lines/words to disk (text)
corpusText <- data.frame(text = unlist(sapply(corpus, '[', "content")), stringsAsFactors = FALSE)
con <- file("data/final/en_US/en_US.corpus.txt", open = "w")
writeLines(corpusText$text, con)
close(con)

kable(head(corpusText$text, 10),
      row.names = FALSE,
      col.names = NULL,
      align = c("l"),
      caption = "First 10 Documents")

# remove variables no longer needed to free up memory
rm(samples, fileStats, corpusText)
gc()
```

### Word Frequencies

A bar chart and word cloud will be constructed to illustrate unique word
frequencies.

```{r word_frequencies, message = FALSE, echo = FALSE}
library(wordcloud)
library(RColorBrewer)

# count words with two or more appearance
tdm <- TermDocumentMatrix(corpus, control = list(bounds = list(local = c(2, Inf))))
freq <- sort(rowSums(as.matrix(tdm)), decreasing = TRUE)
wordFreq <- data.frame(word = names(freq), freq = freq)

# plot the top 10 most frequent words
g <- ggplot (wordFreq[1:10,], aes(x = reorder(wordFreq[1:10,]$word, -wordFreq[1:10,]$fre),
                                  y = wordFreq[1:10,]$fre ))
g <- g + geom_bar( stat = "Identity" , fill = I("grey50"))
g <- g + geom_text(aes(label = wordFreq[1:10,]$fre), vjust = -0.20, size = 3)
g <- g + xlab("")
g <- g + ylab("Word Frequencies")
g <- g + theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(hjust = 0.5, vjust = 0.5, angle = 45),
               axis.text.y = element_text(hjust = 0.5, vjust = 0.5))
g <- g + ggtitle("10 Most Frequent Words")
print(g)

# construct word cloud
suppressWarnings (
    wordcloud(words = wordFreq$word,
              freq = wordFreq$freq,
              min.freq = 1,
              max.words = 100,
              random.order = FALSE,
              rot.per = 0.35, 
              colors=brewer.pal(8, "Dark2"))
)

# remove variables no longer needed to free up memory
rm(tdm, freq, wordFreq, g)
```

### Tokenizing and N-Gram Generation

The predictive model I plan to develop for the Shiny application will handle 
uniqrams, bigrams, and trigrams. In this section, I will use the `RWeka` package
to construct functions that tokenize the sample data and construct matrices of
uniqrams, bigrams, and trigrams.

```{r tokenizers, message = FALSE, echo = FALSE}
library(RWeka)

unigramTokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 1, max = 1))
bigramTokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 2, max = 2))
trigramTokenizer <- function(x) NGramTokenizer(x, Weka_control(min = 3, max = 3))
```

#### Unigrams

```{r unigrams, message = FALSE, echo = FALSE}
# create term document matrix for the corpus
unigramMatrix <- TermDocumentMatrix(corpus, control = list(tokenize = unigramTokenizer))

# eliminate sparse terms for each n-gram and get frequencies of most common n-grams
unigramMatrixFreq <- sort(rowSums(as.matrix(removeSparseTerms(unigramMatrix, 0.99))), decreasing = TRUE)
unigramMatrixFreq <- data.frame(word = names(unigramMatrixFreq), freq = unigramMatrixFreq)

# generate plot
g <- ggplot(unigramMatrixFreq[1:20,], aes(x = reorder(word, -freq), y = freq))
g <- g + geom_bar(stat = "identity", fill = I("grey50"))
g <- g + geom_text(aes(label = freq ), vjust = -0.20, size = 3)
g <- g + xlab("")
g <- g + ylab("Frequency")
g <- g + theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(hjust = 1.0, angle = 45),
               axis.text.y = element_text(hjust = 0.5, vjust = 0.5))
g <- g + ggtitle("20 Most Common Unigrams")
print(g)
```

#### Bigrams

```{r bigrams, message = FALSE, echo = FALSE}
# create term document matrix for the corpus
bigramMatrix <- TermDocumentMatrix(corpus, control = list(tokenize = bigramTokenizer))

# eliminate sparse terms for each n-gram and get frequencies of most common n-grams
bigramMatrixFreq <- sort(rowSums(as.matrix(removeSparseTerms(bigramMatrix, 0.999))), decreasing = TRUE)
bigramMatrixFreq <- data.frame(word = names(bigramMatrixFreq), freq = bigramMatrixFreq)

# generate plot
g <- ggplot(bigramMatrixFreq[1:20,], aes(x = reorder(word, -freq), y = freq))
g <- g + geom_bar(stat = "identity", fill = I("grey50"))
g <- g + geom_text(aes(label = freq ), vjust = -0.20, size = 3)
g <- g + xlab("")
g <- g + ylab("Frequency")
g <- g + theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(hjust = 1.0, angle = 45),
               axis.text.y = element_text(hjust = 0.5, vjust = 0.5))
g <- g + ggtitle("20 Most Common Bigrams")
print(g)
```

#### Trigrams

```{r trigrams, message = FALSE, echo = FALSE}
# create term document matrix for the corpus
trigramMatrix <- TermDocumentMatrix(corpus, control = list(tokenize = trigramTokenizer))

# eliminate sparse terms for each n-gram and get frequencies of most common n-grams
trigramMatrixFreq <- sort(rowSums(as.matrix(removeSparseTerms(trigramMatrix, 0.9999))), decreasing = TRUE)
trigramMatrixFreq <- data.frame(word = names(trigramMatrixFreq), freq = trigramMatrixFreq)

# generate plot
g <- ggplot(trigramMatrixFreq[1:20,], aes(x = reorder(word, -freq), y = freq))
g <- g + geom_bar(stat = "identity", fill = I("grey50"))
g <- g + geom_text(aes(label = freq ), vjust = -0.20, size = 3)
g <- g + xlab("")
g <- g + ylab("Frequency")
g <- g + theme(plot.title = element_text(size = 14, hjust = 0.5, vjust = 0.5),
               axis.text.x = element_text(hjust = 1.0, angle = 45),
               axis.text.y = element_text(hjust = 0.5, vjust = 0.5))
g <- g + ggtitle("20 Most Common Trigrams")
print(g)
```

## Way Forward

The final deliverable in the capstone project is to build a predictive algorithm
that will be deployed as a Shiny app for the user interface. The Shiny app
should take as input a phrase (multiple words) in a text box input and output a
prediction of the next word.

The predictive algorithm will be developed using an n-gram model with a
word frequency lookup similar to that performed in the exploratory data analysis
section of this report. A strategy will be built based on the knowledge 
gathered during the exploratory analysis. For example, as n increased for
each n-gram, the frequency decreased for each of its terms. So one possible
strategy may be to construct the model to first look for the unigram that would
follow from the entered text. Once a full term is entered followed by a space,
find the most common bigram model and so on.

Another possible strategy may be to predict the next word using the trigram
model. If no matching trigram can be found, then the algorithm would check the
bigram model. If still not found, use the unigram model.

The final strategy will be based on the one that increases efficiency and
provides the best accuracy.

## Appendix

### A.1 Basic Data Summary

Basic summary of the three text corpora.

```{r prepare_data-appendix, ref.label = 'prepare_data', echo = TRUE, eval = FALSE}
```

### A.2 Histogram of Words per Line

Histogram of words per line for the three text corpora.

```{r hist_words-appendix, ref.label = 'hist_words', echo = TRUE, eval = FALSE}
```

### A.3 Clean the Data and Build Corpus

```{r clean_data-appendix, ref.label = 'clean_data', echo = TRUE, eval = FALSE}
```

### A.4 Word Frequencies

```{r word-frequencies-appendix, ref.label = 'word_frequencies', echo = TRUE, eval = FALSE}
```

### A.5 Tokenizing and N-Gram Generation

**Tokenize Functions**

```{r tokenize-appendix, ref.label = 'tokenizers', echo = TRUE, eval = FALSE}
```

**Unigrams**

```{r unigrams-appendix, ref.label = 'unigrams', echo = TRUE, eval = FALSE}
```

**Bigrams**

```{r bigrams-appendix, ref.label = 'bigrams', echo = TRUE, eval = FALSE}
```

**Trigrams**

```{r trigrams-appendix, ref.label = 'trigrams', echo = TRUE, eval = FALSE}
```
